\MakeUppercase{Graph Product Structure for Non-Minor-Closed Classes}

### 2022-03-17 response.tex

\begin{itemize}
\item graphs drawn on a fixed surface with each edge in $\leq k$ crossings,
\item map graphs of degree $\leq k$ graphs embedded in a fixed surface,
\item string graphs with each curve in $\leq k$ intersections,
\item $k$-nearest neighbour graphs in the plane, and
\item powers of classes of graphs of degree $\leq k$ which admit a product structure theorem.
\end{itemize}

All but the last bullet point are shown to follow from the first.
The main results in the paper are correct, strong, and of high interest,
and the paper should definitely be accepted to JCTB after some
(fairly significant) revisions. In particular, the proof of Theorem 3 is
difficult to read due to the notation, and the paper could be substantially
streamlined without losing the above results. Detailed suggestions follow.

Major comments:

% The following comments (and several additional minor comments) are all related to the proof of Theorem~3 in the original submission.  Much of the difficulty comes from the fact that there are many things to keep track of. We are a given a graph $G$, a shortcut system $\mathcal{S}$, a layering $\mathcal{L}$, a vertex-partition $\mathcal{P}$, and a tree decomposition $\mathcal{T}$ of the quotient graph $H:=G/\mathcal{P}$.  Now we have to use these to show the existence of a new layering $\mathcal{L}'$, a new vertex partition $\mathcal{P'}$, a new tree decomposition $\mathcal{T}'$ of the new quotient graph $H':=G^{\mathcal{S}}$.
%
% During the proof, a number of related sets come up, which we have given names to. As a referee point out, these could remain unnamed and be described using an appropriate notation.
%
% \todo[inline]{Write something more here depending on how the experiment with new notation works out.}
%
% One issue is that a number of the proofs in this section are essentially "definition checking".  The relationships between the various vertex subsets that arise in the proof really do just follow (by design) from their definitions.  In the paper, we do this checking explicitly, though each of these proofs could also be replaced by a sentence of the form ``This [claim/lemma] follows from the definitions of\ldots''.
%
% However, when discussing the proof of Theorem~3 with colleagues (very familiar with product structure) recently, we found them unconvinced by high-level explanations.  It was only after carefully unpacking the definitions that they became convinced.  These conversations usually end with someone saying ``Ah, of course!''
%
%
% The real action begins in the proof Theorem 9 (using numbering from the original submission) and especially Claim 3 (which involves some careful counting) and Claim 5 (which involves checking that we have everything needed to apply the Pillipczuk-Siebertz lemma on $k$-reachability in ``upward-directed'' graphs of treewidth $t$.)
%
% \todo[inline]{Write more here depending on how the notation experiment turns out.}

 % and new vertex partition of a related graph $G^\mathcal{S}$ as we  A large part of the

@@

1. Lemma 3: This lemma is really about ``normalised'' tree decompositions, and says that for each $x \in V(T)$, the set $V(T_x)$ has at most $t$ neighbours in $H$. Call this property (T3) and remove (Y1)  through (Y5), which should now be obvious.

This lemma has been eliminated completely.

2. Proof of Theorem 9: There is too much notation throughout. As
one example, it is hard to remember which of $T_x$, $Y_x$, $V_x$, $F_x$, $N_x$,
$S_x$, $X_v$, $B_x$, $C_x$ is which. If $\phi(S)$ is written for the subset of $V(G)$
corresponding to a set $S \subseteq V(H)$ and appropriate shorthand and
standard notation for neighbourhoods are used, then $Y_x$, $V_x$, $F_x$, and
$N_x$ can be replaced by $\phi(x)$, $\phi(T_x)$, $N(\phi(T_x))$, and $V(G) \setminus N[\phi(T_x)]$.
The set $X_v$ is only used in the statement of Claim~1, so it does not
need a name; say what these vertices are explicitly. It may be more
indicative to write $\phi'(x)$ for $S_x$ and $B'_x$ for $C_x$.

We agree that the notation was excessive (an unfortunate consequence of the general result on $(k,d)$-shortcut systems having evolved from a specific result about $k$-planar graphs).  We have eliminated the notations $V_x$, $N_x$, $F_x$, $X_x$, and $T_x$.

The notations $Y_x$ and $B_x$ are needed since we need names for the parts of the $H$-decomposition of $G$ and the bags in the tree-decomposition of $H$.    The notation $C_x$ is limited to the proof of (newly-numbered) Claim~3, where it is used to denote the bags in the tree-decomposition that is the subject of Claim~3.

We have avoided the introduction of any new notation.  The new presentation is much shorter and clearer.  We thank the referee for making us revisit this proof carefully.  The final paper is better for it.

@@

3. Theorem 10: Unfortunately I have to suggest removing it. I don’t see
the motivation for making $H$ planar or for improving the constants.
Are the improved constants at all close to tight? Is there another
reason for including it?

Since the initial submission, there is now even more justification for including this special case.  A result of Bose et al on vertex ranking relies critically on the simple treewidth of $H$.  In particular, if $H$ is planar and has treewidth $3$, then it has an $\ell$-vertex ranking using $O(\log n/\log\log\log n)$ colours.  The three logs in the denominator come from the fact

\todo[inline]{Continue here...}

### 2022-03-17 k-planar.tex
Added osme justification
Specified that $T_0$ has no empty bags

Consider a tree $T$ rooted at some node $x_0\in V(T)$. Each subtree $T'$ of $T$ is considered to be rooted at the node in $T'$ closest to $x_0$. A node $a\in V(T)$ is a \defin{$T$-ancestor} of $x\in V(T)$ (and $x$ is a \defin{$T$-descendant} of $a$) if $a$ is a vertex of the path, in $T$, from $x_0$ to $x$.  Note that each node $x\in V(T)$ is a $T$-ancestor and $T$-descendant of itself.  We say that a $T$-ancestor $a\in V(T)$ of $x\in V(T)$ is a \defin{strict} $T$-ancestor of $x$ if $a\neq x$.
The \defin{$T$-depth} of a node $x\in V(T)$ is the length of the path, in $T$, from $x_0$ to $x$.  For each node $x\in V(T)$, define
\[T_x := T[\{y\in V(T):\mbox{$x$ is a $T$-ancestor of $y$}\}] \]
to be the maximal subtree of $T$ rooted at $x$.

@@

  Begin with any width-$t$ tree-decomposition $(B_x:x\in V(T_0))$ of $H$ that uses some tree $T_0$.  Select any node $x\in V(T_0)$, add a leaf $x_0$, with $B_{x_0}=\emptyset$, adjacent to $x$ and root $T_0$ at $x_0$. (The purpose of $x_0$ is to ensure that every node $x$ for which $B_x$ is non-empty has a parent.)  Let $f:V(H)\to V(T)$ be the function that maps each $x\in V(H)$ onto the root of the subtree $T_0[x]:=T_0[\{y\in V(T_0): x\in B_y\}]$.  If $f$ is not one-to-one, then select some distinct pair $x,y\in V(H)$ with $a:=f(x)=f(y)$.  Subdivide the edge between $a$ and its parent in $T_0$ by introducing a new node $a'$ with $B_{a'}=B_{a}\setminus\{x\}$. Now $f(y)=a'$ and $f(x)=a$, so this modification reduces the number of distinct pairs $x,y\in V(H)$ with $f(x)=f(y)$, so repeatedly performing this modification will eventually produce a tree-decomposition $(B_x:x\in V(T_0))$ of $H$ in which $f$ is one-to-one.

>>>>

Consider a tree $T$ rooted at some node $x_0\in V(T)$. Each subtree $T'$ of $T$ is considered to be rooted at the node in $T'$ closest to $x_0$. A node $a\in V(T)$ is a \defin{$T$-ancestor} of $x\in V(T)$ (and $x$ is a \defin{$T$-descendant} of $a$) if $a$ is a vertex of the path, in $T$, from $x_0$ to $x$.  Note that each node $x\in V(T)$ is a $T$-ancestor and $T$-descendant of itself.  We say that a $T$-ancestor $a\in V(T)$ of $x\in V(T)$ is a \defin{strict} $T$-ancestor of $x$ if $a\neq x$.
The \defin{$T$-depth} of a node $x\in V(T)$ is the length of the path, in $T$, from $x_0$ to $x$.
% For each node $x\in V(T)$, define
% \[T_x := T[\{y\in V(T):\mbox{$x$ is a $T$-ancestor of $y$}\}] \]
% to be the maximal subtree of $T$ rooted at $x$.

@@

  Begin with any width-$t$ tree-decomposition $(B_x:x\in V(T_0))$ of $H$ that uses some tree $T_0$ and has no empty bags.  Select any node $x\in V(T_0)$, add a leaf $x_0$, with $B_{x_0}=\emptyset$, adjacent to $x$ and root $T_0$ at $x_0$. (The purpose of $x_0$ is to ensure that every node $x$ for which $B_x$ is non-empty has a parent.)  Let $f:V(H)\to V(T)$ be the function that maps each $x\in V(H)$ onto the root of the subtree $T_0[x]:=T_0[\{y\in V(T_0): x\in B_y\}]$.  If $f$ is not one-to-one, then select some distinct pair $x,y\in V(H)$ with $a:=f(x)=f(y)$.  Subdivide the edge between $a$ and its parent in $T_0$ by introducing a new node $a'$ with $B_{a'}=B_{a}\setminus\{x\}$. Now $f(y)=a'$ and $f(x)=a$, so this modification reduces the number of distinct pairs $x,y\in V(H)$ with $f(x)=f(y)$, so repeatedly performing this modification will eventually produce a tree-decomposition $(B_x:x\in V(T_0))$ of $H$ in which $f$ is one-to-one.
